---
title: 'Exp01: Pilot Soybean Stomatal Conductance Analysis'
author: "Jua'nele Webb"
date: "2026-06-24"
output: html_document
---

## Overview

### Goal

Pilot test of whether leaf age (position on the stem) affects how
much stomatal conductance drops under drought, comparing one droughted
soybean plant to one well-watered control.

### Inputs

`experiment_01.csv`, with columns `Date`, `Leaf_Position`,
`Plant_ID` (`Soybean01_drought` / `Soybean02_control`), and
`Stomatal_Conductance_mmol_m2_s`. Measurements were taken with an SC-1
porometer on 6/11/2026 and 6/15/2026 across multiple leaf positions per
plant (higher position number = younger leaf).

### Design Note

This is a pilot/proof-of-concept run with n = 1 plant per
treatment group (one drought, one well-watered), not a replicated
experiment. Treat results as hypothesis-generating rather than conclusive
(see Conclusion below).

### Outputs

1. A bar chart of percent reduction in conductance (drought vs. control) by
   leaf position on 6/15/2026, with the underlying values and summary stats.
2. A line/point chart comparing conductance across leaf positions and dates
   for both plants, with treatment-and-date means.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(ggplot2)

# Load raw stomatal conductance data
data <- read.csv("experiment_01.csv")
```

## Experiment Window: Gas Exchange vs Leaf Position (06/11/2026-06/15/2026)

### 1. Percent Reduction Relative to Well-Watered Control

Compares the drought soybean plant to the well-watered control on
6/15/2026, using:

Percent Reduction = ((Control − Drought) / Control) × 100

#### Code

```{r percent-reduction-615}
# Reshape data and compute percent reduction
comparison_615 <- data %>%
  filter(Date == "6/15/2026") %>%
  select(
    Leaf_Position,
    Plant_ID,
    Stomatal_Conductance_mmol_m2_s
  ) %>%
  pivot_wider(
    names_from = Plant_ID,
    values_from = Stomatal_Conductance_mmol_m2_s
  ) %>%
  mutate(
    Percent_Reduction =
      ((Soybean02_control - Soybean01_drought) /
         Soybean02_control) * 100
  )

# Plot percent reduction by leaf position
ggplot(
  comparison_615,
  aes(
    x = factor(Leaf_Position),
    y = Percent_Reduction
  )
) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Percent Reduction Relative to Well-Watered Control on 6/15/2026",
    x = "Leaf Position",
    y = "Percent Reduction (%)"
  ) +
  theme_minimal()
```

#### Values Used in Graph

```{r percent-reduction-table}
# Print percent reduction values per leaf position
comparison_615 %>%
  select(Leaf_Position, Percent_Reduction) %>%
  print(n = Inf)
```

#### Mean Percent Reduction Statistics

```{r percent-reduction-summary}
# Compute mean and SD of reduction
comparison_615 %>%
  summarise(
    Mean_Reduction = mean(Percent_Reduction, na.rm = TRUE),
    SD = sd(Percent_Reduction, na.rm = TRUE)
  )
```

#### Interpretation

The average reduction in stomatal conductance across leaf positions was
38.6% relative to the well-watered control. The magnitude of reduction
varied among leaf positions (SD = 23.7%). One leaf position (leaf 4)
deviated from the overall trend; however, the corresponding control leaf
displayed visible disease symptoms, which may have influenced the
measurement. The magnitude of reduction varied with leaf position, ranging
from approximately 17% to 63%, suggesting that different-aged leaves
responded differently to water limitation. The strongest reductions
occurred in leaf positions 6–10, where conductance was approximately
50–63% lower than the control. Lower leaf positions generally exhibited
smaller reductions.

### 2. Comparison of Stomatal Conductance Across Leaf Positions

Compares stomatal conductance of the drought-treated and well-watered
soybean plants across leaf positions on 6/11/2026 and 6/15/2026. Each point
represents an individual plant.

#### Code

```{r comparison-graph}
# Plot conductance across leaf positions and dates
ggplot(
  data,
  aes(
    x = Leaf_Position,
    y = Stomatal_Conductance_mmol_m2_s,
    color = Plant_ID,
    group = Plant_ID
  )
) +
  geom_point(size = 3) +
  geom_line() +
  facet_wrap(~Date) +
  scale_color_manual(
    name = "Treatment",
    values = c(
      "Soybean01_drought" = "lightblue",
      "Soybean02_control" = "steelblue"
    ),
    labels = c(
      "Drought",
      "Well-watered Control"
    )
  ) +
  labs(
    title = "Comparing Stomatal Conductance Across Leaf Positions in Droughted and Well-Watered Soybeans",
    subtitle = "Higher leaf position numbers correspond to younger leaves",
    x = "Leaf Position",
    y = "Stomatal Conductance (mmol m⁻² s⁻¹)"
  )
```

#### Mean Stomatal Conductance by Treatment and Date

```{r sc-by-treatment-date}
# Summarize mean conductance by date and plant
data %>%
  group_by(Date, Plant_ID) %>%
  summarise(
    Mean = mean(Stomatal_Conductance_mmol_m2_s, na.rm = TRUE),
    SD = sd(Stomatal_Conductance_mmol_m2_s, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Leaf Positions with the Largest Drought-Induced Reduction

```{r max-reduction}
# Rank leaf positions by reduction size
comparison_615 %>%
  arrange(desc(Percent_Reduction))
```

#### Interpretation

Mean stomatal conductance decreased from 69.4 to 56.4 mmol m⁻² s⁻¹ in the
drought-treated plant between sampling dates. Mean stomatal conductance
increased from 52.6 to 97.0 mmol m⁻² s⁻¹ in the well-watered control.

## Conclusion

Tool used: SC-1.

Pilot experiment has perceived relativity to proposal's anticipated result:
younger leaves may maintain higher stomatal conductivity under drought
stress. Leading to the result: leaf age may influence stomatal conductance
and further WUE, but data set was too small (one droughted, one watered,
one conductivity measurement) and leaf variation was too wide to be
certain. A clear age-related trend was difficult to note. Additional
replication needed.

## Limitations / Next Steps

- n = 1 plant per treatment group — no statistical test is valid at this
  sample size; percent-reduction values should be read as descriptive only.
- Leaf 4's control showed visible disease symptoms, a likely confound for
  that data point specifically.
- Needs replication (more plants per treatment) before drawing conclusions
  about leaf-age effects on drought response, which is picked up in
  `exp02_sc_tepary.Rmd` onward.
