---
title: "exp02: Tepary Stomatal Conductance"
author: "Jua'nele Webb"
date: "2026-07-07"
output: html_document
---

## Overview

### Goal

Test whether stomatal conductance (SC) changes with meristem leaf
position (M1–M3) differently across drought vs. watered tepary bean plants,
and whether that relationship shifts across the Baseline → Treatment →
Recovery timeline.

### Inputs

`exp02_tepary_SC1_20260625.csv` — one row per plant/date/leaf
position, with meristem-position SC readings for 4 plants (Plant 1–4).
Column 6 is stomatal conductance and is renamed `SC` for readability.

### Design

2 erect plants (Plant 1, Plant 2) and 2 climbing plants
(Plant 3, Plant 4); within each architecture, one plant was droughted and
one kept as a watered control (Plant 1 & 3 = drought, Plant 2 & 4 =
control). Measurements span 6/25–7/2/2026:

- **Baseline:** 6/25, 6/26
- **Treatment (drought imposed):** 6/29
- **Recovery:** 7/1, 7/2

### Outputs

1. A table of the 4 plants' architectural traits.
2. `meristem_stage_plot` — SC vs. meristem position, faceted by architecture
   and date, colored by experimental stage.
3. A per-date, per-treatment regression of SC on meristem position (slope,
   R², p-value, MSE), plus a scatter/trend view of the same relationship.
4. R² and MSE trend lines over time, marking when drought begins.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
library(knitr)

# Load raw meristem SC data
data <- read.csv("exp02_tepary_SC1_20260625.csv")

# Quick inspection of the raw import
names(data)
head(data)
dim(data)
unique(data$Position)
```

## 1. Plant Architecture

```{r architecture-table}
# Build architecture summary table
architecture <- data.frame(
  Trait = c(
    "Growth Habit", "Height", "Architecture", "Internodes",
    "Branching", "Canopy", "Apical Dominance",
    "Shoot Tip", "Lateral Growth",
    "Support Needed"
  ),
  `Plant 1` = c(
    "Erect", "Short", "Compact", "Short",
    "High", "Compact", "Reduced",
    "Damaged", "High", "No"
  ),
  `Plant 2` = c(
    "Erect", "Medium", "Upright", "Long",
    "Low", "Open", "Strong",
    "Intact", "Low", "No"
  ),
  `Plant 3` = c(
    "Climber", "Tall", "Vine-like", "Long",
    "Moderate", "Open", "Reduced",
    "Damaged", "High", "Yes"
  ),
  `Plant 4` = c(
    "Climber", "Tall", "Vine-like", "Long",
    "Moderate", "Moderate", "Strong",
    "Intact", "Moderate", "Yes"
  ),
  check.names = FALSE  # keeps "Plant 1".."Plant 4" as the column headers
                        # instead of R silently renaming them "Plant.1" etc.
)

kable(
  architecture,
  caption = "Architectural characteristics of four tepary bean plants."
)
```

- **Plant 1:** Compact, erect, highly branched.
- **Plant 2:** Upright with few branches and strong apical dominance.
- **Plant 3:** Climbing plant with a damaged shoot tip and active lateral growth.
- **Plant 4:** Climbing plant with an intact shoot tip and moderate branching.

## 2. Data Cleaning & Labeling

```{r data-cleaning}
# Rename the raw stomatal-conductance column for readability
names(data)[6] <- "SC"

# Derive architecture, stage, treatment, and position labels
data_clean <- data %>%
  mutate(
    Date = factor(
      Date,
      levels = c(
        "6/25/2026", "6/26/2026", "6/29/2026",
        "7/1/2026", "7/2/2026"
      )  # chronological order, needed below for the trend-line plots
    ),

    SC = as.numeric(SC),

    Architecture = case_when(
      Plant_ID %in% c("Plant 1", "Plant 2") ~ "Erect",
      Plant_ID %in% c("Plant 3", "Plant 4") ~ "Climber"
    ),

    # 3-phase timeline, used for the stage-colored meristem plot
    Stage = case_when(
      Date %in% c("6/25/2026", "6/26/2026") ~ "Baseline",
      Date == "6/29/2026" ~ "Treatment",
      Date %in% c("7/1/2026", "7/2/2026") ~ "Recovery"
    ),
    Stage = factor(Stage, levels = c("Baseline", "Treatment", "Recovery")),

    # Baseline/Drought/Control grouping, used for the regression-by-treatment
    # statistics — baseline dates are their own group regardless of plant,
    # since drought hadn't been imposed yet.
    Treatment = case_when(
      Stage == "Baseline" ~ "Baseline",
      Plant_ID %in% c("Plant 1", "Plant 3") ~ "Drought",
      Plant_ID %in% c("Plant 2", "Plant 4") ~ "Control"
    ),
    Treatment = factor(Treatment, levels = c("Baseline", "Drought", "Control")),

    Experimental_Group = case_when(
      Plant_ID %in% c("Plant 1", "Plant 3") ~ "Drought",
      Plant_ID %in% c("Plant 2", "Plant 4") ~ "Watered Control"
    ),
    Experimental_Group = factor(
      Experimental_Group,
      levels = c("Watered Control", "Drought")
    ),

    Leaf_Type = ifelse(grepl("^M", Position), "Meristem", "Lateral"),
    Meristem_Position = case_when(
      Position == "M1" ~ 1,
      Position == "M2" ~ 2,
      Position == "M3" ~ 3,
      TRUE ~ NA_real_
    )
  )

# Check assignments
data_clean %>%
  distinct(Date, Plant_ID, Architecture, Experimental_Group, Stage, Treatment) %>%
  arrange(Date, Plant_ID)

# Filter to meristem-position rows only
meristem_data <- data_clean %>%
  filter(
    Leaf_Type == "Meristem",
    !is.na(Stage),
    !is.na(Treatment)
  )
```

## 3. Meristem Position and Stomatal Conductance Across Experimental Stages

```{r meristem-stage-plot, fig.width=9, fig.height=6}
# Plot SC by position, stage, architecture
meristem_stage_plot <- ggplot(
  meristem_data,
  aes(
    x = Meristem_Position,
    y = SC,
    color = Stage,
    shape = Experimental_Group,
    linetype = Experimental_Group,
    group = interaction(Plant_ID, Date)
  )
) +
  geom_point(size = 3, alpha = 0.85) +
  geom_smooth(
    aes(group = interaction(Stage, Experimental_Group, Architecture, Date)),
    method = "lm",
    se = FALSE,
    linewidth = 1
  ) +
  facet_grid(
    Architecture ~ Date,
    labeller = labeller(Date = function(x) format(as.Date(x, "%m/%d/%Y"), "%m/%d"))
  ) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("M1", "M2", "M3")) +
  scale_color_manual(
    values = c(
      "Baseline" = "#7570B3",
      "Treatment" = "#D95F02",
      "Recovery" = "#1B9E77"
    )
  ) +
  labs(
    title = "Meristem Position and Stomatal Conductance Across Experimental Stages",
    subtitle = "Baseline: 6/25-6/26 | Treatment: 6/29 | Recovery: 7/1-7/2",
    x = "Meristem Leaf Position",
    y = expression("Stomatal Conductance (mmol " * m^-2 * " " * s^-1 * ")"),
    color = "Experimental Stage",
    shape = "Water Treatment",
    linetype = "Water Treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

meristem_stage_plot
```

## 4. Regression of SC on Meristem Position, by Date and Treatment

For each date x treatment group, fits `SC ~ Meristem_Position` and reports
the slope, R², p-value, and mean squared error (MSE).

```{r regression-stats}
# Fit regression per date and treatment
stats_table <- meristem_data %>%
  group_by(Date, Treatment) %>%
  group_modify(~{
    fit <- lm(SC ~ Meristem_Position, data = .x)
    glance_out <- glance(fit)
    coef_out <- tidy(fit)

    tibble(
      Slope = coef_out$estimate[2],
      R2 = glance_out$r.squared,
      P_value = coef_out$p.value[2],
      MSE = mean(residuals(fit)^2)
    )
  }) %>%
  ungroup()

stats_table
```

```{r regression-scatter}
# Plot SC vs position by treatment
ggplot(
  meristem_data,
  aes(x = Meristem_Position, y = SC, color = Treatment)
) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~Date, labeller = labeller(Date = function(x) format(as.Date(x, "%m/%d/%Y"), "%m/%d"))) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("M1", "M2", "M3")) +
  labs(
    title = "Relationship Between Meristem Leaf Position and Stomatal Conductance",
    subtitle = "Plants grouped by drought treatment",
    x = "Meristem Leaf Position",
    y = expression("Stomatal Conductance (mmol " * m^-2 * s^-1 * ")")
  ) +
  theme_classic()
```

## 5. R² and MSE Trends Over Time

```{r r2-mse-trends}
# Find x-position where drought begins
drought_start_pos <- which(levels(stats_table$Date) == "6/29/2026")

# Plot R-squared trend over time
ggplot(
  stats_table,
  aes(x = Date, y = R2, color = Treatment, group = Treatment)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = drought_start_pos, linetype = "dashed") +
  annotate(
    "text",
    x = drought_start_pos,
    y = max(stats_table$R2, na.rm = TRUE),
    label = "Drought begins",
    vjust = -0.5,
    size = 3
  ) +
  labs(
    title = "Change in R² Over Time",
    subtitle = "R² shows how much stomatal conductance variation is explained by meristem position",
    x = "Sampling Date",
    y = expression(R^2),
    color = "Condition"
  ) +
  theme_minimal()

# Plot MSE trend over time
ggplot(
  stats_table,
  aes(x = Date, y = MSE, color = Treatment, group = Treatment)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = drought_start_pos, linetype = "dashed") +
  annotate(
    "text",
    x = drought_start_pos,
    y = max(stats_table$MSE, na.rm = TRUE),
    label = "Drought begins",
    vjust = -0.5,
    size = 3
  ) +
  labs(
    title = "Change in MSE Over Time",
    subtitle = "MSE shows regression error between observed and predicted stomatal conductance",
    x = "Sampling Date",
    y = "Mean Squared Error",
    color = "Condition"
  ) +
  theme_minimal()
```

## Limitations / Next Steps

- Only one plant per architecture x treatment combination — as with Exp01,
  treat regression results (slope/R²/p-value) as descriptive per plant
  rather than as inferential statistics across a population.
