---
title: 'exp04: Stomatal Conductance Analysis for Compact and Erect Soybeans'
author: "Jua'nele Webb"
date: "2026-08-02"
output:
  html_document: default
  pdf_document: default
---

## Overview

### Goal

Test whether decapitating the shoot apical meristem (forcing a
more compact architecture) changes how soybean stomatal conductance
responds to drought over time, compared to intact ("Erect") plants —
looking at the highest and lowest mature meristem positions separately.

### Inputs

`exp04_cut_soybeans_sc1_20260729.csv` — one row per plant x
date, with `Cut_Status` (Cut = decapitated/"Compact", Intact = "Erect"),
`Water_Status` (Watered/Drought), `Day` (Day 1-3), and separate SC columns
for the lowest and highest mature meristem (`Lowest_Mature_Meristem_SC`,
`Highest_Mature_Meristem_SC`), pivoted here into one `SC` column tagged by
`Mature_Position`.

### Outputs

1. `trajectory_plot` — mean SC over sampling day, by architecture, faceted
   by water status x mature position, with a per-panel ANOVA p-value for
   the architecture effect. Saved to `trajectory_plot_NPG.png` / `.pdf`.
2. `drought_penalty_plot` — drought-vs-watered SC difference over time, by
   architecture and mature position.
3. `stability_plot` — coefficient of variation in SC across dates, as a
   measure of temporal stability.
4. `architecture_position_plot` — interaction between architecture and
   mature position.
5. `redistribution_plot` plus a mixed-effects model (`redistribution_model`)
   of how SC is redistributed between the highest and lowest mature
   positions over time.

### Required Packages

Beyond the tidyverse: `broom`, `scales`, `ggsci`,
`lme4`, `lmerTest`. Install once with
`install.packages(c("broom","scales","ggsci","lme4","lmerTest"))` if any
`library()` call below fails.

## 1. Import and Clean

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(broom)
library(ggsci)
library(lme4)
library(lmerTest)

# Load raw cut-soybean SC data
data <- read.csv(
  "exp04_cut_soybeans_sc1_20260729.csv",
  check.names = FALSE
)

# Inspect the original dataset
names(data)
head(data)
str(data)

# Standardize architecture, water status, and day
data_clean <- data %>%
  mutate(
    Date = as.Date(Date, format = "%m/%d/%Y"),

    Architecture = case_when(
      Cut_Status == "Cut" ~ "Compact",
      Cut_Status == "Intact" ~ "Erect",
      TRUE ~ NA_character_
    ),
    Architecture = factor(Architecture, levels = c("Erect", "Compact")),

    Water_Status = factor(Water_Status, levels = c("Watered", "Drought")),
    Plant_ID = factor(Plant_ID),

    Day = factor(Day, levels = c("Day 1", "Day 2", "Day 3"))
  )

# One row per plant, date, position
analysis_data <- data_clean %>%
  pivot_longer(
    cols = c(Lowest_Mature_Meristem_SC, Highest_Mature_Meristem_SC),
    names_to = "Mature_Position",
    values_to = "SC"
  ) %>%
  mutate(
    Mature_Position = recode(
      Mature_Position,
      "Lowest_Mature_Meristem_SC" = "Lowest Mature",
      "Highest_Mature_Meristem_SC" = "Highest Mature"
    ),
    Mature_Position = factor(Mature_Position, levels = c("Lowest Mature", "Highest Mature")),
    SC = as.numeric(gsub("[^0-9.-]", "", SC))
  )

# Confirm 4a and 4b remain
analysis_data %>%
  filter(Plant_ID %in% c("4a", "4b")) %>%
  arrange(Plant_ID, Date, Mature_Position)

# Check sample sizes
analysis_data %>%
  count(Date, Architecture, Water_Status, Mature_Position)

# Check for missing values
analysis_data %>%
  filter(is.na(SC))
```

## 2. Trajectory Analysis

Mean SC over sampling day, by architecture, faceted by water status x
mature position, testing the architecture effect within each panel.

```{r trajectory-summary}
# Summarize mean SC by day and architecture
trajectory_summary <- analysis_data %>%
  group_by(Day, Architecture, Water_Status, Mature_Position) %>%
  summarise(
    n = sum(!is.na(SC)),
    Mean_SC = mean(SC, na.rm = TRUE),
    SD_SC = sd(SC, na.rm = TRUE),
    SE_SC = SD_SC / sqrt(n),
    .groups = "drop"
  )

trajectory_summary
```

```{r trajectory-anova}
# Test architecture effect within each panel
trajectory_anova <- analysis_data %>%
  filter(!is.na(SC), !is.na(Architecture), !is.na(Day), !is.na(Water_Status), !is.na(Mature_Position)) %>%
  group_by(Water_Status, Mature_Position) %>%
  group_modify(~{
    panel_data <- droplevels(.x)
    panel_model <- aov(SC ~ Architecture * Day, data = panel_data)
    model_results <- broom::tidy(panel_model)
    architecture_result <- model_results %>% filter(term == "Architecture")
    data.frame(Architecture_p = architecture_result$p.value)
  }) %>%
  ungroup()

trajectory_anova
```

```{r trajectory-plot, fig.width=8, fig.height=6}
# Find max SC per panel for labels
panel_maximums <- trajectory_summary %>%
  group_by(Water_Status, Mature_Position) %>%
  summarise(Panel_Maximum = max(Mean_SC + SE_SC, na.rm = TRUE), .groups = "drop")

# Build ANOVA label position and text
trajectory_anova_labels <- trajectory_anova %>%
  left_join(panel_maximums, by = c("Water_Status", "Mature_Position")) %>%
  mutate(
    ANOVA_Label = paste0("ANOVA, p = ", format.pval(Architecture_p, digits = 3, eps = 0.001)),
    x_position = 1,
    y_position = Panel_Maximum + 0.12 * max(trajectory_summary$Mean_SC + trajectory_summary$SE_SC, na.rm = TRUE)
  )

trajectory_anova_labels

# Plot SC trajectory by architecture
trajectory_plot <- ggplot(
  trajectory_summary,
  aes(x = Day, y = Mean_SC, color = Architecture, group = Architecture)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Mean_SC - SE_SC, ymax = Mean_SC + SE_SC), width = 0.08, linewidth = 0.7) +
  geom_text(
    data = trajectory_anova_labels,
    aes(x = x_position, y = y_position, label = ANOVA_Label),
    inherit.aes = FALSE, hjust = 0, size = 3.6
  ) +
  facet_grid(Water_Status ~ Mature_Position) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
  labs(
    x = "Sampling Day",
    y = expression("Stomatal Conductance (mmol " * m^-2 * " " * s^-1 * ")"),
    color = "Architecture"
  ) +
  theme_classic(base_size = 12) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10, color = "black"),
    legend.position = "bottom",
    legend.title = element_text(size = 9, face = "bold"),
    legend.text = element_text(size = 8),
    panel.spacing = unit(0.8, "lines")
  )

trajectory_plot

# Apply journal color palette
trajectory_plot_NPG <- trajectory_plot +
  scale_color_npg() +
  theme(legend.position = "bottom", legend.title = element_text(face = "bold"), legend.text = element_text(size = 9))

trajectory_plot_NPG

# Save trajectory plot as PNG and PDF
ggsave("trajectory_plot_NPG.png", trajectory_plot_NPG, width = 8, height = 6, units = "in", dpi = 600, bg = "white")
ggsave("trajectory_plot_NPG.pdf", trajectory_plot_NPG, width = 8, height = 6, units = "in", device = cairo_pdf, bg = "white")
```

### Spot check: Watered x Lowest Mature panel

`trajectory_anova` above already fits `SC ~ Architecture * Day` for every
Water Status x Mature Position panel — this just prints the full model
summary for one panel (Watered, Lowest Mature) as a sanity check on that
number.

```{r spot-check-panel}
# Filter to one panel for spot check
panel_check <- analysis_data %>%
  filter(Water_Status == "Watered", Mature_Position == "Lowest Mature") %>%
  droplevels()

table(panel_check$Architecture, panel_check$Day)

# Fit architecture x day model
panel_model <- aov(SC ~ Architecture * Day, data = panel_check)
summary(panel_model)
```

### Interpretation

This graph shows whether compact and erect plants maintained stomatal
conductance differently over time. It also distinguishes watered and
droughted plants and the highest and lowest mature positions.

## 3. Drought Penalty

```{r drought-penalty, fig.width=8, fig.height=5}
# Summarize mean SC by treatment and date
penalty_summary <- analysis_data %>%
  group_by(Date, Architecture, Mature_Position, Water_Status) %>%
  summarise(
    n = sum(!is.na(SC)),
    Mean_SC = mean(SC, na.rm = TRUE),
    SD_SC = sd(SC, na.rm = TRUE),
    SE_SC = SD_SC / sqrt(n),
    .groups = "drop"
  )

# Compute drought penalty vs watered mean
drought_penalty_data <- penalty_summary %>%
  select(Date, Architecture, Mature_Position, Water_Status, Mean_SC, SE_SC) %>%
  pivot_wider(names_from = Water_Status, values_from = c(Mean_SC, SE_SC)) %>%
  mutate(
    Drought_Penalty = Mean_SC_Drought - Mean_SC_Watered,
    Penalty_SE = sqrt(SE_SC_Drought^2 + SE_SC_Watered^2)
  )

drought_penalty_data

# Plot drought penalty over time
drought_penalty_plot <- ggplot(
  drought_penalty_data,
  aes(
    x = Date, y = Drought_Penalty, color = Architecture,
    linetype = Mature_Position, shape = Mature_Position,
    group = interaction(Architecture, Mature_Position)
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.7) +
  geom_errorbar(aes(ymin = Drought_Penalty - Penalty_SE, ymax = Drought_Penalty + Penalty_SE), width = 0.08, linewidth = 0.6, alpha = 0.75) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.2) +
  scale_x_date(date_labels = "%m/%d", date_breaks = "1 day") +
  labs(
    title = "Architecture Effects on Drought Penalty Across Shoot Positions",
    subtitle = "Drought penalty = drought mean - corresponding watered mean",
    x = "Date",
    y = expression(Delta * " Mean Stomatal Conductance (Drought - Watered)"),
    color = "Architecture",
    linetype = "Mature Position",
    shape = "Mature Position"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )

drought_penalty_plot
```

## 4. Stability

```{r stability, fig.width=8, fig.height=5}
# Compute coefficient of variation per plant
stability_data <- analysis_data %>%
  group_by(Plant_ID, Architecture, Water_Status, Mature_Position) %>%
  summarise(
    n_dates = sum(!is.na(SC)),
    Mean_SC = mean(SC, na.rm = TRUE),
    SD_SC = sd(SC, na.rm = TRUE),
    CV_SC = SD_SC / Mean_SC,
    SC_Range = max(SC, na.rm = TRUE) - min(SC, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_dates >= 2, is.finite(CV_SC))

# Summarize CV by architecture and treatment
stability_summary <- stability_data %>%
  group_by(Architecture, Water_Status, Mature_Position) %>%
  summarise(n = n(), Mean_CV = mean(CV_SC, na.rm = TRUE), SD_CV = sd(CV_SC, na.rm = TRUE), SE_CV = SD_CV / sqrt(n), .groups = "drop")

stability_summary

# Plot CV by architecture and treatment
stability_plot <- ggplot(
  stability_data,
  aes(x = Architecture, y = CV_SC, fill = Water_Status)
) +
  geom_boxplot(position = position_dodge(width = 0.75), width = 0.62, alpha = 0.65, outlier.shape = NA) +
  geom_point(
    aes(shape = Mature_Position),
    position = position_jitterdodge(jitter.width = 0.10, dodge.width = 0.75),
    size = 2.8, alpha = 0.85
  ) +
  labs(
    title = "Temporal Stability of Stomatal Conductance by Architecture",
    subtitle = "Lower coefficients of variation indicate greater stability across dates",
    x = "Architecture",
    y = "Coefficient of Variation Across Dates",
    fill = "Water Status",
    shape = "Mature Position"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )

stability_plot
```

### Interpretation

Lower variability indicates more stable conductance over time. This should
be interpreted with the trajectory graph because a consistently low
conductance value can also produce low variability.

## 5. Architecture x Mature-Position Interaction

```{r interaction-plot, fig.width=9, fig.height=5}
# Summarize mean SC by position and date
interaction_summary <- analysis_data %>%
  group_by(Date, Architecture, Water_Status, Mature_Position) %>%
  summarise(
    n = sum(!is.na(SC)),
    Mean_SC = mean(SC, na.rm = TRUE),
    SD_SC = sd(SC, na.rm = TRUE),
    SE_SC = SD_SC / sqrt(n),
    .groups = "drop"
  )

# Plot architecture x position interaction
architecture_position_plot <- ggplot(
  interaction_summary,
  aes(
    x = Mature_Position, y = Mean_SC, color = Architecture,
    linetype = Water_Status, shape = Water_Status,
    group = interaction(Architecture, Water_Status)
  )
) +
  geom_line(linewidth = 1.1, position = position_dodge(width = 0.10)) +
  geom_point(size = 3.2, position = position_dodge(width = 0.10)) +
  geom_errorbar(aes(ymin = Mean_SC - SE_SC, ymax = Mean_SC + SE_SC), width = 0.08, linewidth = 0.7, position = position_dodge(width = 0.10)) +
  facet_wrap(~Date, nrow = 1, labeller = labeller(Date = function(x) format(as.Date(x), "%m/%d"))) +
  labs(
    title = "Interaction Between Shoot Architecture and Mature Position",
    subtitle = "Nonparallel lines suggest architecture affects shoot positions differently",
    x = "Mature Meristem Position",
    y = expression("Mean Stomatal Conductance (mmol " * m^-2 * " " * s^-1 * ")"),
    color = "Architecture",
    linetype = "Water Status",
    shape = "Water Status"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    strip.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )

architecture_position_plot
```

## 6. Developmental Redistribution Between Mature Positions

```{r redistribution, fig.width=8, fig.height=5}
# Compute SC difference between positions
redistribution_data <- data_clean %>%
  mutate(Delta_SC = Highest_Mature_Meristem_SC - Lowest_Mature_Meristem_SC)

# Summarize redistribution by architecture and date
redistribution_summary <- redistribution_data %>%
  group_by(Date, Architecture, Water_Status) %>%
  summarise(n = n(), Mean_Delta = mean(Delta_SC, na.rm = TRUE), SD_Delta = sd(Delta_SC, na.rm = TRUE), SE_Delta = SD_Delta / sqrt(n), .groups = "drop")

redistribution_summary

# Plot redistribution trend over time
redistribution_plot <- ggplot(
  redistribution_summary,
  aes(x = Date, y = Mean_Delta, color = Architecture, group = Architecture)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.7) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Mean_Delta - SE_Delta, ymax = Mean_Delta + SE_Delta), width = 0.08) +
  facet_wrap(~Water_Status) +
  scale_x_date(date_labels = "%m/%d", date_breaks = "1 day") +
  labs(
    title = "Developmental Redistribution of Stomatal Conductance",
    subtitle = "Positive values indicate greater conductance in the highest mature meristem",
    x = "Date",
    y = expression(Delta * " Stomatal Conductance (Highest - Lowest)"),
    color = "Architecture"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

redistribution_plot
```

```{r redistribution-model}
# Fit mixed model of redistribution over time
redistribution_model <- lmer(
  Delta_SC ~ Architecture * Water_Status * Date + (1 | Plant_ID),
  data = redistribution_data
)

anova(redistribution_model)
```

## Limitations / Next Steps

- Sample size per Architecture x Water Status combination is small (a
  handful of plants, e.g. "4a"/"4b" sub-replicates) — treat panel-specific
  ANOVA p-values as suggestive rather than confirmatory.
- `redistribution_model` treats `Date` as a continuous predictor in a
  3-way interaction with a random intercept for `Plant_ID`; with only 3
  sampling days this estimates a linear trend over time rather than
  day-specific effects — worth revisiting if more timepoints are added.
