---
title: "exp03: Hormone Effects on Leaf Age Patterns in Stomatal Conductance"
author: "Jua'nele Webb"
date: "2026-07-23"
output: html_document
---

## Research Question

Do IAA and paclobutrazol treatments alter the developmental distribution of
stomatal conductance between the youngest and oldest leaves of tepary bean
under well-watered and drought conditions?

## Biological Question

The analysis evaluates whether hormone treatments alter:

1. overall stomatal conductance;
2. developmental differences between youngest and oldest leaves;
3. relative physiological performance under drought;
4. the distribution of conductance toward younger or older tissues.

## Hypotheses

1. Youngest leaves will maintain greater stomatal conductance than oldest
   leaves under drought.
2. IAA may increase physiological support of younger leaves under drought.
3. PAC may alter stomatal conductance through growth restriction and more
   conservative water use.
4. Hormone treatments may change the magnitude or direction of
   developmental differences in stomatal conductance.

## Overview

### Inputs

`exp03_hormone_tepary_SC1_20260716.csv`, one row per plant x
date x leaf, with `Treatment` (Control/IAA/PAC), `Water_Status`
(Well-watered/Drought), `Leaf_Age` (Oldest/Youngest), `Leaf_Status`
(only "Healthy" rows are analyzed), and column 6 = stomatal conductance
(renamed `SC`).

### Outputs

For each of five derived metrics — developmental gradient,
developmental prioritization, drought retention, developmental plasticity,
drought response, and developmental allocation — a summary table, a plot,
and (where applicable) an ANOVA / Tukey HSD test.

### Required Packages

Beyond the tidyverse: `broom`, `patchwork`,
`ggpubr`, `multcompView`, `ggsci`, `scales`. Install once with
`install.packages(c("broom","patchwork","ggpubr","multcompView","ggsci","scales"))`
if any `library()` call below fails — installation is left out of the
knitted chunks so re-running this notebook doesn't try to reinstall
packages every time.

## Setup, Import, and Clean

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(patchwork)
library(scales)
library(multcompView)
library(ggpubr)
library(ggsci)

# Load raw hormone SC data
data <- read.csv("exp03_hormone_tepary_SC1_20260716.csv")

names(data)[6] <- "SC"

# Standardize factor levels and clean SC
data <- data %>%
  mutate(
    Treatment = factor(Treatment, levels = c("Control", "IAA", "PAC")),
    Water_Status = factor(Water_Status, levels = c("Well-watered", "Drought")),
    Leaf_Age = factor(Leaf_Age, levels = c("Oldest", "Youngest")),
    Leaf_Age = trimws(as.character(Leaf_Age)),
    SC = as.numeric(gsub("[^0-9.-]", "", SC))
  )
```

```{r clean-and-gradient}
# Keep only healthy leaf measurements
healthy_data <- data %>%
  filter(Leaf_Status == "Healthy")

# Pair youngest/oldest SC within each plant x date x treatment and compute the developmental gradient (Youngest - Oldest). Positive values mean higher conductance in the youngest leaf; negative means higher in the oldest leaf.
gradient_data <- healthy_data %>%
  select(Date, Plant_ID, Treatment, Water_Status, Leaf_Age, SC) %>%
  pivot_wider(names_from = Leaf_Age, values_from = SC) %>%
  mutate(Developmental_Gradient = Youngest - Oldest) %>%
  drop_na(Oldest, Youngest, Developmental_Gradient)

gradient_data
```

### Interpretation

The data were cleaned by converting stomatal conductance to numeric values,
standardizing leaf-age labels, and retaining only healthy leaves. Youngest
and oldest leaf measurements were paired within each plant to calculate the
developmental gradient:

Youngest − Oldest stomatal conductance

Positive values indicate higher conductance in the youngest leaf, while
negative values indicate higher conductance in the oldest leaf.

## Developmental Gradient by Hormone

```{r gradient-by-hormone, fig.width=9, fig.height=5}
# Subset to well-watered plants
control_gradient <- gradient_data %>% filter(Water_Status == "Well-watered")

# Summarize gradient stats for control
control_summary <- control_gradient %>%
  group_by(Treatment) %>%
  summarise(
    n = n(),
    Mean_Gradient = mean(Developmental_Gradient, na.rm = TRUE),
    SD = sd(Developmental_Gradient, na.rm = TRUE),
    SE = SD / sqrt(n),
    Median = median(Developmental_Gradient),
    Minimum = min(Developmental_Gradient),
    Maximum = max(Developmental_Gradient),
    .groups = "drop"
  )
control_summary

# Plot gradient by treatment, well-watered
control_plot <- ggplot(
  control_gradient,
  aes(x = Treatment, y = Developmental_Gradient, fill = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.7) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.65) +
  geom_jitter(width = 0.10, size = 2.7) +
  labs(
    title = "Well-Watered",
    x = "Hormone Treatment",
    y = expression(Delta * " Stomatal Conductance\n(Youngest - Oldest)"),
    fill = "Treatment"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    plot.margin = margin(8, 10, 8, 8)
  )

# Test treatment effect, well-watered group
control_model <- aov(Developmental_Gradient ~ Treatment, data = control_gradient)
summary(control_model)
TukeyHSD(control_model)

# Subset to drought plants
drought_gradient <- gradient_data %>% filter(Water_Status == "Drought")

# Summarize gradient stats for drought
drought_summary <- drought_gradient %>%
  group_by(Treatment) %>%
  summarise(
    n = n(),
    Mean_Gradient = mean(Developmental_Gradient, na.rm = TRUE),
    SD = sd(Developmental_Gradient, na.rm = TRUE),
    SE = SD / sqrt(n),
    Median = median(Developmental_Gradient, na.rm = TRUE),
    Minimum = min(Developmental_Gradient, na.rm = TRUE),
    Maximum = max(Developmental_Gradient, na.rm = TRUE),
    .groups = "drop"
  )
drought_summary

# Plot gradient by treatment, drought
drought_plot <- ggplot(
  drought_gradient,
  aes(x = Treatment, y = Developmental_Gradient, fill = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.7) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.65) +
  geom_jitter(width = 0.10, size = 2.7) +
  labs(title = "Drought", x = "Hormone Treatment", y = NULL, fill = "Treatment") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    plot.margin = margin(8, 8, 8, 10)
  )

# Test treatment effect, drought group
drought_model <- aov(Developmental_Gradient ~ Treatment, data = drought_gradient)
summary(drought_model)
TukeyHSD(drought_model)

# Combine control and drought plots
combined_gradient_plot <-
  (control_plot | drought_plot) +
  plot_layout(guides = "collect", widths = c(1, 1)) +
  plot_annotation(
    title = "Developmental Gradient by Hormone Treatment in Tepary",
    theme = theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 12, hjust = 0.5)
    )
  ) &
  theme(legend.position = "bottom")

combined_gradient_plot
```

### Interpretation

IAA showed the greatest positive gradient under well-watered conditions,
while PAC showed the greatest positive gradient under drought. Due to
PAC's inhibition of gibberellin biosynthesis, the maintenance of gas
exchange throughout the stress treatment potentially infers greater water
conservation.

PAC observed with the reduced climber architecture demonstrated from
PhenoSight RGB.

## Developmental Prioritization, Relative Drought Performance, Developmental Plasticity

```{r prioritization}
# Normalized developmental prioritization index
prioritization_data <- gradient_data %>%
  mutate(Developmental_Prioritization = (Youngest - Oldest) / (Youngest + Oldest))

# Print prioritization values per plant
prioritization_data %>%
  select(Plant_ID, Treatment, Water_Status, Oldest, Youngest, Developmental_Prioritization) %>%
  print(n = Inf)

# Summarize prioritization by treatment and water
prioritization_summary <- prioritization_data %>%
  group_by(Water_Status, Treatment) %>%
  summarise(
    n = n(),
    Mean_Prioritization = mean(Developmental_Prioritization, na.rm = TRUE),
    SD = sd(Developmental_Prioritization, na.rm = TRUE),
    SE = SD / sqrt(n),
    .groups = "drop"
  )
prioritization_summary

# Plot prioritization index by treatment
prioritization_plot <- ggplot(
  prioritization_data,
  aes(x = Treatment, y = Developmental_Prioritization, fill = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.7) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA, width = 0.60) +
  geom_jitter(width = 0.10, size = 2.8) +
  facet_wrap(~Water_Status) +
  labs(
    title = "Developmental Prioritization of Stomatal Conductance",
    x = "Hormone Treatment",
    y = "Developmental Prioritization Index"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "none"
  )
prioritization_plot

# Test treatment x water interaction
prioritization_model <- aov(Developmental_Prioritization ~ Treatment * Water_Status, data = prioritization_data)
summary(prioritization_model)
```

```{r drought-retention}
# Compute drought retention percent by leaf age
retention_data <- healthy_data %>%
  group_by(Treatment, Leaf_Age, Water_Status) %>%
  summarise(Mean_SC = mean(SC, na.rm = TRUE), SD_SC = sd(SC, na.rm = TRUE), n = n(), .groups = "drop") %>%
  select(Treatment, Leaf_Age, Water_Status, Mean_SC) %>%
  pivot_wider(names_from = Water_Status, values_from = Mean_SC) %>%
  mutate(Drought_Retention_Percent = Drought / `Well-watered` * 100)

retention_data

# Plot drought retention by treatment
retention_plot <- ggplot(
  retention_data,
  aes(x = Treatment, y = Drought_Retention_Percent, fill = Leaf_Age)
) +
  geom_hline(yintercept = 100, linetype = "dashed", linewidth = 0.7) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  labs(
    title = "Relative Drought Performance of Stomatal Conductance",
    subtitle = "100% indicates equal mean stomatal conductance under drought and well-watered conditions",
    x = "Hormone Treatment",
    y = "Mean Drought Stomatal Conductance (% of Well-Watered)",
    fill = "Leaf Age"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10),
    legend.position = "bottom"
  )
retention_plot
```

```{r plasticity-data}
# Compute absolute and relative plasticity
plasticity_data <- gradient_data %>%
  mutate(
    Developmental_Plasticity = abs(Youngest - Oldest),
    Relative_Plasticity = abs(Youngest - Oldest) / ((Youngest + Oldest) / 2)
  )

# Summarize plasticity by water and treatment
plasticity_summary <- plasticity_data %>%
  group_by(Water_Status, Treatment) %>%
  summarise(
    n = n(),
    Mean_Absolute_Plasticity = mean(Developmental_Plasticity, na.rm = TRUE),
    SD_Absolute_Plasticity = sd(Developmental_Plasticity, na.rm = TRUE),
    Mean_Relative_Plasticity = mean(Relative_Plasticity, na.rm = TRUE),
    SD_Relative_Plasticity = sd(Relative_Plasticity, na.rm = TRUE),
    .groups = "drop"
  )
plasticity_summary

# Test treatment x water interaction
plasticity_model <- aov(Relative_Plasticity ~ Treatment * Water_Status, data = plasticity_data)
summary(plasticity_model)
```

Now that we have ANOVA values, Tukey HSD shows which groups are actually
significantly different from each other, split by water status:

```{r plasticity-tukey}
# Split plasticity data by water status
control_pd <- subset(plasticity_data, Water_Status == "Well-watered")
drought_pd <- subset(plasticity_data, Water_Status == "Drought")

# Helper: get Tukey letters per group
tukey_letters_for <- function(df, water_label) {
  fit <- TukeyHSD(aov(Relative_Plasticity ~ Treatment, data = df))
  p_adj <- fit$Treatment[, "p adj"]
  letters_out <- multcompLetters(p_adj)
  out <- as.data.frame(letters_out$Letters)
  out$Treatment <- rownames(out)
  colnames(out)[1] <- "Tukey"
  out$Water_Status <- water_label
  out
}

# Compute Tukey letters for both groups
test_all <- rbind(
  tukey_letters_for(control_pd, "Well-watered"),
  tukey_letters_for(drought_pd, "Drought")
)
test_all

# Reorder water status factor levels
plasticity_data$Water_Status <- factor(plasticity_data$Water_Status, levels = c("Well-watered", "Drought"))
test_all$Water_Status <- factor(test_all$Water_Status, levels = c("Well-watered", "Drought"))
```

```{r plasticity-plot, fig.width=8, fig.height=5}
# Plot plasticity with significance labels
plasticity_plot <- ggplot(
  plasticity_data,
  aes(x = Treatment, y = Relative_Plasticity, fill = Treatment)
) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA, width = 0.60) +
  geom_jitter(width = 0.10, size = 2.8) +
  facet_wrap(~Water_Status) +
  labs(
    title = "Developmental Plasticity in Stomatal Conductance",
    x = "Hormone Treatment",
    y = "Relative Plasticity ((Y-O)/mean(Y,O))"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "none"
  ) +
  stat_compare_means(method = "anova", label.y = 1) +
  stat_pvalue_manual(test_all, label = "Tukey", y.position = 0.85)

plasticity_plot

# Apply journal color palette
plasticity_plot_NPG <- plasticity_plot + scale_fill_aaas()
plasticity_plot_NPG
```

### Interpretation

#### Developmental Prioritization

PAC: highest mean prioritization index (0.142)
IAA: slightly negative prioritization index (-0.039)

#### Relative Drought Performance

All treatments: reduced stomatal conductance under drought (<100%
retention)
IAA: greatest conductance in older leaves (78.7%)
PAC: retained greatest conductance in the youngest leaf (61.7%)

#### Developmental Plasticity

PAC: highest relative plasticity (0.467)
IAA: lowest relative plasticity (0.077)

## Hormone Stomatal Conductance Response

```{r drought-response, fig.width=8, fig.height=5}
# Compute SC change under drought by age
drought_response_data <- healthy_data %>%
  group_by(Treatment, Leaf_Age, Water_Status) %>%
  summarise(Mean_SC = mean(SC, na.rm = TRUE), SD_SC = sd(SC, na.rm = TRUE), n = n(), .groups = "drop") %>%
  select(Treatment, Leaf_Age, Water_Status, Mean_SC) %>%
  pivot_wider(names_from = Water_Status, values_from = Mean_SC) %>%
  mutate(
    Drought_Response = Drought - `Well-watered`,
    Absolute_Decline = `Well-watered` - Drought
  )

drought_response_data

# Plot drought response by treatment and age
drought_response_plot <- ggplot(
  drought_response_data,
  aes(x = Treatment, y = Drought_Response, fill = Leaf_Age)
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.7) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  labs(
    title = "Hormone-Specific Stomatal Conductance Response to Drought",
    x = "Hormone Treatment",
    y = expression(Delta * " Mean Stomatal Conductance (Drought - Well-Watered)"),
    fill = "Leaf Age"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
drought_response_plot

# Rank treatments by drought response
drought_response_ranking <- drought_response_data %>%
  arrange(Leaf_Age, desc(Drought_Response)) %>%
  group_by(Leaf_Age) %>%
  mutate(Response_Rank = row_number()) %>%
  ungroup()

drought_response_ranking
```

### Interpretation

IAA: smallest decline in oldest (-6.62 mmol/m⁻² s⁻¹), but largest decline
in youngest (-36.92 mmol/m⁻² s⁻¹)
PAC: more balanced response across ages (-17.99 to -20.84 mmol/m⁻² s⁻¹)
Control: more intermediate (-23.08 to -27.28 mmol/m⁻² s⁻¹)

## Water Use Allocation Index

```{r allocation-index, fig.width=9, fig.height=5}
# Compute youngest-leaf share of conductance
allocation_data <- gradient_data %>%
  mutate(Developmental_Allocation_Index = Youngest / (Youngest + Oldest))

# Print allocation index per plant
allocation_data %>%
  select(Plant_ID, Treatment, Water_Status, Oldest, Youngest, Developmental_Allocation_Index) %>%
  print(n = Inf)

# Summarize allocation by water and treatment
allocation_summary <- allocation_data %>%
  group_by(Water_Status, Treatment) %>%
  summarise(
    n = n(),
    Mean_Allocation = mean(Developmental_Allocation_Index, na.rm = TRUE),
    SD_Allocation = sd(Developmental_Allocation_Index, na.rm = TRUE),
    SE_Allocation = SD_Allocation / sqrt(n),
    Median_Allocation = median(Developmental_Allocation_Index, na.rm = TRUE),
    .groups = "drop"
  )
allocation_summary

# Overall interaction model
allocation_model <- aov(Developmental_Allocation_Index ~ Treatment * Water_Status, data = allocation_data)
summary(allocation_model)

# Test ANOVA and Tukey by water status
allocation_well_watered_model <- aov(
  Developmental_Allocation_Index ~ Treatment,
  data = allocation_data %>% filter(Water_Status == "Well-watered")
)
summary(allocation_well_watered_model)
TukeyHSD(allocation_well_watered_model)

allocation_drought_model <- aov(
  Developmental_Allocation_Index ~ Treatment,
  data = allocation_data %>% filter(Water_Status == "Drought")
)
summary(allocation_drought_model)
TukeyHSD(allocation_drought_model)

# Helper: compute Tukey letters for one group
get_allocation_letters <- function(df) {
  model <- aov(Developmental_Allocation_Index ~ Treatment, data = df)
  tukey_results <- TukeyHSD(model)
  adjusted_p_values <- tukey_results$Treatment[, "p adj"]
  letter_results <- multcompLetters(adjusted_p_values)$Letters

  data.frame(Treatment = names(letter_results), Tukey_Letter = unname(letter_results))
}

# Get Tukey letters per water status
allocation_letters <- allocation_data %>%
  group_by(Water_Status) %>%
  group_modify(~ get_allocation_letters(.x)) %>%
  ungroup() %>%
  mutate(
    Treatment = factor(Treatment, levels = c("Control", "IAA", "PAC")),
    Water_Status = factor(Water_Status, levels = c("Well-watered", "Drought"))
  )
allocation_letters

# Compute label y-position per group
allocation_letter_positions <- allocation_data %>%
  group_by(Water_Status, Treatment) %>%
  summarise(Letter_Position = max(Developmental_Allocation_Index, na.rm = TRUE) + 0.035, .groups = "drop") %>%
  left_join(allocation_letters, by = c("Water_Status", "Treatment"))
allocation_letter_positions

# Compute ANOVA p-value label per panel
allocation_anova_labels <- allocation_data %>%
  group_by(Water_Status) %>%
  group_modify(~{
    panel_model <- aov(Developmental_Allocation_Index ~ Treatment, data = .x)
    panel_p_value <- summary(panel_model)[[1]]["Treatment", "Pr(>F)"]
    data.frame(ANOVA_Label = paste0("ANOVA, p = ", format.pval(panel_p_value, digits = 2, eps = 0.001)))
  }) %>%
  ungroup() %>%
  mutate(
    Water_Status = factor(Water_Status, levels = c("Well-watered", "Drought")),
    x_position = 1,
    y_position = 0.86
  )
allocation_anova_labels

# Plot allocation index with significance labels
allocation_plot <- ggplot(
  allocation_data,
  aes(x = Treatment, y = Developmental_Allocation_Index, fill = Treatment)
) +
  geom_hline(yintercept = 0.50, linetype = "dashed", linewidth = 0.7) +
  geom_boxplot(alpha = 0.65, outlier.shape = NA, width = 0.60) +
  geom_jitter(width = 0.10, size = 2.8) +
  facet_wrap(~Water_Status) +
  geom_text(
    data = allocation_letter_positions,
    aes(x = Treatment, y = Letter_Position, label = Tukey_Letter),
    inherit.aes = FALSE, size = 5
  ) +
  geom_text(
    data = allocation_anova_labels,
    aes(x = x_position, y = y_position, label = ANOVA_Label),
    inherit.aes = FALSE, hjust = 0, size = 4.5
  ) +
  labs(
    title = "Developmental Allocation of Stomatal Conductance",
    x = "Hormone Treatment",
    y = "Youngest Leaf Share of Combined Conductance"
  ) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    breaks = seq(0.4, 0.9, by = 0.1),
    limits = c(0.4, 0.9),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  coord_cartesian(clip = "off") +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "none",
    plot.margin = margin(10, 15, 10, 10)
  )

allocation_plot

# Apply journal color palette
allocation_plot_NPG <- allocation_plot + scale_fill_aaas()
allocation_plot_NPG
```

### Interpretation

Allocation Index measures stomatal conductance distributed between older
and younger leaves. Values above 50% indicate greater allocation to
youngest leaf, while values below 50% indicate greater allocation to older
leaves.

PAC: maintained highest allocation to the youngest leaf with 57.1% mean
allocation
IAA: shifted towards greater allocation in older leaves with 48.1% mean
allocation

## Limitations / Next Steps

- No pre-treatment baseline or post-drought recovery measurements — drought
  and well-watered plants are independent groups, not repeated measures on
  the same plants.
